### 2021-10-06 typesetting/sex_selection_spacing-figure_typesetting.tex
Figures for typesetting at Demography

% For word counting uncomment line below and raggedright below
% http://countwordsfree.com/#
% \usepackage[nolists]{endfloat}

% \bibpunct{(}{)}{;}{a}{}{,}
\newcommand{\mco}[1]{\multicolumn{1}{c}{#1}}
\newcommand{\mct}[1]{\multicolumn{2}{c}{#1}}
\newcommand{\mcth}[1]{\multicolumn{3}{c}{#1}}
\newcommand{\X}{$\times$ }
\newcommand{\hs}{\hspace{15pt}}

% Attempt to squeeze more floats in
\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
Fourth Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
Text and code for this paper is available at 
\href{https://github.com/cportner/sexSelectionSpacing}{https://github.com/cportner/sexSelectionSpacing}
}
}

@@

\begin{document}
\graphicspath{{../figures/}}
\DeclareGraphicsExtensions{.eps,.jpg,.pdf,.mps,.png}

@@

% \setcounter{page}{0}

@@

\begin{abstract}

\noindent Following the arrival of prenatal sex-determination technologies in the mid-1980s, 
India has experienced an increasingly male-biased sex ratio at birth, presumably from sex-selective 
abortions. 
Abortions lengthen birth intervals, but we know little about how birth spacing has changed or the 
effects of these changes. 
This paper shows that, although the length of birth intervals increased overall, well-educated women 
with no sons had the most substantial lengthening---and the most male-biased sex ratios. 
Furthermore, most of these changes took place immediately after the introduction of prenatal 
sex-determination technologies. 
Consequently, some women without sons now have longer birth intervals than those with sons, reversing 
India's traditional spacing pattern. 
Less-educated women continue short birth spacing when they have no sons, with only limited evidence 
of male-biased sex ratios. 
Because of the rapid lengthening of birth intervals, period fertility rates substantially overestimated 
how fast cohort fertility fell. 
Moreover, predicted cohort fertility is still 10\%--20\% above the period fertility rate. 
If lengthening of birth intervals arise from repeated abortions, the associated short pregnancy spacing 
may counteract any positive effects of longer birth spacing. 
There is, however, no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.

\noindent JEL: J1, O12, I1

\noindent Keywords: India, prenatal sex determination, censoring, competing risk, 
nonproportional hazard
\end{abstract}

@@

\section{Introduction\label{sec:intro}}

% For word counting remove comment on the line below
% \raggedright

% Background
India has experienced many positive changes over the past four decades:
The economy has grown substantially,
real wages have more than doubled,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,Klasen2015,
International-Institute-for-Population-Sciences-IIPS2017}.

However, India also witnessed the advent of sex selection---the selective abortion of 
female fetuses based on prenatal sex determination---with the introduction of ultrasound 
in the mid-1980s.
Combined with a strong, continued preference for sons, the result was a dramatic increase 
in the male--female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the use of sex-selective abortions could 
substantially increase birth spacing. 
As detailed below, after an abortion it takes 6 months or more to reach the same point in 
the next pregnancy. 
Second, other major societal changes, such as women's greater educational attainment, higher 
household income, and a low and declining female labor force participation, all 
likely influence birth spacing. 
The combination of the apparent increasing use of sex selection and societal 
changes can significantly impact birth spacing, but we know little about how much.

Studying birth spacing contributes to our understanding of fertility decisions, but, equally 
important, birth spacing also affects the reliability of our fertility measures and may 
affect mortality. 
Therefore, I address two additional questions.

First, did changes in birth spacing lead us to overestimate the decline in cohort fertility?
With longer spacing, mothers will be older at each parity, and this tempo effect 
makes period fertility measures, such as the total fertility rate, downward-biased
estimates of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, a rapidly expanding use of sex selection could make the total fertility rate---our 
most used fertility measure---fall substantially faster than cohort fertility.
In this case, households' fertility may be higher than generally accepted.

Second, what is the relationship between infant mortality and the changes in birth spacing
and sex selection? 
In India, birth intervals have traditionally been shorter with fewer sons, contributing to 
girls' higher mortality risk
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}. 
Therefore, longer birth spacing may reduce mortality through, for example, diminished 
sibling competition \citep{Conde-Agudelo2012,Molitoris2019}. 
However, if the spacing between births lengthens because of sex-selective abortions, the 
spacing between pregnancies may still be very short. 
Short pregnancy spacing may lead to worse child outcomes because of maternal nutritional 
depletion and insufficient time to recover from the previous pregnancy. 
Hence, children born after long birth intervals where multiple abortions have punctuated 
those intervals may not see the same benefits as children born after a long interval that 
is not punctuated by abortions.

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: The birth of a girl or the birth of a boy. 
I apply the model to Hindu women's birth histories covering 1972--2016 using data from the 
four National Family and Health Surveys (NFHS).

The primary outcomes I examine are the 25th, 50th, and 75th percentile birth intervals 
durations;
the sex ratio at birth; and the likelihood of giving birth. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key explanatory variables are maternal education, the sex of previous children, and 
the area of residence.

The empirical model allows me to predict cohort fertility. 
To examine whether tempo effects bias our standard fertility measures, I compare 
the predicted cohort fertility with fertility calculated from age-specific fertility 
rates.

I use the same data to study how infant mortality changed with birth spacing and 
the increasing use of sex selection. 
The key explanatory variables remain the same, except for the addition of birth spacing 
and the sex of the index child.

% Main results
There are three main results.

First, birth intervals lengthened over the four decades, with the lengthening longer, the
higher the parity, the more educated the woman, and the higher the percentile of the
birth interval duration.
Well-educated women with no sons had the most substantial lengthening of birth intervals 
and the most male-biased sex ratios, both likely arising from sex-selective abortions. 
Consequently, some women with no sons now have \emph{longer} birth intervals than those 
with sons, reversing India's traditional spacing pattern. 
Less-educated women continued to have short spacing when they have girls, with only limited 
evidence of male-biased sex ratios.
The likelihood of a very short birth interval changed little across most groups.

Second, the period fertility rate substantially overestimated how fast cohort fertility 
fell in the 1990s and early 2000s as spacing initially increased. 
Although the two have lately been converging, the predicted cohort fertility is still 
10\%--20\% higher than the period fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

Finally, infant mortality has declined substantially over time for all groups, but fastest 
for the less educated, who are now close to the level of the best-educated women. 
However, mortality is still inversely related to education level, especially for very short 
birth intervals. 
There is no evidence that repeated sex-selective abortions are associated with higher 
mortality for the child eventually born.

%------------------------------------------------------------------------------------

\section{Conceptual Framework}

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, investment in children, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education, area of residence, and sex composition are the
principal factors in the empirical analyses and tie them to the three explanations.

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility, although the direction is ambiguous; empirically, 
higher female wages reduce fertility, whereas higher male wages increase fertility 
\citep{Hotz1997,schultz97}.
According to economic theory, the substitution and income effects' relative strengths 
determine the effect of higher wages on fertility.
The substitute effect captures that when wages increase, time becomes more expensive, and
people, therefore, work more and spend less time on non-wage earning activities, such as
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Because women spend substantially more time on childrearing than men, the substitution
effect dominates for women while the income effect dominates for men.

Higher female wages may also shorten birth spacing if having children requires mothers to
curtail their economic activities.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children requires less than double the time 
needed for one child \citep{Vijverberg1982,Hotz1997}.

% Investment in children
With increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means a stronger incentive to invest in their children's
education, thereby increasing the cost of having children and lowering fertility. 
With lower expected mortality, parents need fewer births to reach the desired number of
surviving children, which, in turn, allows parents to invest more in each child.

An increased desire to invest in children may also lengthen birth spacing. 
The clearest example of a positive effect of longer spacing is for health, as very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Stronger son preference has traditionally been associated with shorter spacing and worse
health outcomes for daughters when there no sons in the household
\citep{Whitworth2002,Jayachandran2011}.
Although son preference increases average fertility only marginally, differential stopping
behavior means that girls tend to end up in larger families, resulting in fewer resources
per child \citep{repetto72,clark00,Basu2010,Barcellos2014}.

However, with the advent of sex selection, stronger son preference may lead to longer
spacing because each sex-selective abortion adds 6–12 months to the length of the birth
interval.
After an abortion, the uterus needs at least two menstrual cycles---approximately 
two months---to recover, or the subsequent risk of spontaneous abortion increases 
substantially \citep{zhou00b}.
Once conception can be attempted, the waiting time to conception is likely between
one and six months. 
Finally, sex-determination tests are reliable only from three months of gestation. 
Hence, it takes at least six months before the couple is at the same point in the next 
pregnancy as they were in the prior pregnancy when they decided to abort.

% Factors/predictions
The closest available proxies in the data for capturing the explanations linking
fertility and birth spacing are the sex composition of previous children, mothers'
education, and area of residence. 
The remainder of the section discusses how the three variables fit into the
explanations and the predicted effects.

Sex composition proxies for son preference and the use of
sex selection, both unobserved.%
\footnote{
The surveys ask about the ideal number of sons and 
daughters, but these appear to be unreliable son 
preference indicators; better-educated women simultaneously show 
declining son preference and increasing sex selection 
\citep{bhat03,pande07}.
} 
In the absence of sons, sex selection appears to increase 
with lower desired fertility and higher parity
\citep{Portner2015b,Jayachandran2017}. 
Thus, for a given parity, birth intervals should lengthen 
substantially over time for women with no sons, lower 
desired fertility, and better access to sex selection.

Both higher female education and urbanization are associated with 
lower fertility and increased sex selection use, and, therefore, likely also changes in birth spacing
\citep{das_gupta97,retherford03b,Portner2015b}.
Furthermore, both variables are representative of
the tremendous changes in India. 
Figure \ref{fig:education_over_time} shows the substantial increase in female education,
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older
using the four rounds of the NFHS}
\label{fig:education_over_time}
\end{figure}

Urban women are likely to have lower fertility and higher use of sex selection than women 
in rural areas.
The lower fertility results from the higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, children in Indian slums have worse
health outcomes than children in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from
higher income and better access to health care facilities in urban areas.

Even though higher potential wages from more education should shorten birth spacing and
lower fertility, the low and declining female labor force participation suggests little
economic incentive to space children closer together.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.

The decline in female labor force participation appears to be driven by a combination
of a relatively small expansion in the sectors where women work and the income effect 
from rapidly increasing male wages and education dominating the substitution effect from 
higher female wages \citep{Klasen2015,Bhargava2018}.
The income effect can dominate because not only do men have a higher average level of 
education than women, men are also paid more than women for a given level of education, 
and the returns to education increases with education in India \citep{Agrawal2011}.
Hence, increasing male education will result in a larger increase in income than a 
similar increase in female education.

However, higher female education might still be associated with lower fertility and more 
sex selection, provided better-educated women are better at producing child human capital, 
as suggested by \citet{Behrman1999}. 
In a situation with increasing returns to male education, households may desire 
better-educated women because of their ability to produce better-educated sons, 
rather than because of their higher potential income. 
The higher return to male education provides an additional incentive to invest in sons' 
education---with a better-educated mother one such investment---and leads to lower desired 
fertility, makes sex selection more likely, and possibly increases birth spacing.
The implication is that the usual policy recommendation of increased female education may not
change sex selection unless there is a concurrent increase in the relative return to female
education in the labor market.

% Sanskritization
Finally, the process of "Sanskritization" implies that as lower-caste females gain access
to education, they adopt higher-caste norms such as stronger son preference and a
retraction from the formal labor market \citep{Srinivas1956}. 
The declining female labor force participation suggests that this process still 
operates \citep{Abraham2013,Chatterjee2018}.
Thus, although the expansion in female education will likely bring in groups who initially 
have different norms, we are unlikely to see substantial changes in how women with a given 
education level behave.

% Summary / hypotheses
In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated as their
household income increase the most, and because of their presumptive use of sex selection. 
Finally, "Sanskritization" implies that the changing composition of better-educated
women will not substantially change the use of sex selection.

\section{Estimation Strategy\label{sec:strategy}}

% Requirements for method and the intuition behind each:  
%  - Account for censoring (hazard model)
%  - Work both with and without sex selection (competing risk part)
%  - Allow for changes in use of sex selection within spell (flexible specification of baseline hazard ?)
%  - Capture that the shape of the hazard function differ across groups
%    (nonproportionality); needs a better description, maybe using example
%    of shorter spacing and/or use of sex selection. Differences in parity
%    progression likelihood.

The empirical analysis has three parts. 
First, I document the changes in birth spacing over time and how the introduction of sex 
selection influenced birth spacing. 
Second, I compare period fertility estimates with predicted cohort fertility based on the 
model of birth spacing. 
Finally, I examine how birth intervals affect infant mortality and whether the apparent 
increase in sex selection influenced mortality. 
This section presents the empirical model for my analysis of birth spacing.

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit---the birth of a child---but there are two problems with the 
standard approach in this setting.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}

First and foremost, the use of sex selection means that the sex of the next child is 
no longer random and that the spacing to the birth of a boy will differ from the 
spacing to a girl. 
Therefore, I use a competing risk setup that captures both the non-randomness of 
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there was underreporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, even without sex selection, it is unlikely that characteristics, such as previous
births' sex composition, have the same effects throughout the entire birth interval.
The proportional hazard model requires that an individual's hazard is a 
fixed proportion of the hazard for any other individual. 
Nonconstant effects violate that assumption, and the results from a proportional hazard 
model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals because there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008}.

The introduction of prenatal sex determination exacerbates any bias from
the proportionality assumption for two reasons. 
First, different groups have different levels of sex-selective abortion use and, thereby, 
birth spacing. 
Second, within a birth interval, a household's use of sex selection may vary, and that
means that the effects of covariates vary as well.

Therefore, I use a nonproportional hazard specification that allows the shape of the 
hazard functions to vary across groups.
Furthermore, the combination of a nonproportional specification and a flexible baseline
hazard mitigates the potential effects of unobserved heterogeneity \citep{Dolton1995}.

The model is a discrete-time, nonproportional, competing risk hazard model with two 
exit states: Either a boy or a girl is born. 
The unit of analysis is a spell---the period from one parity birth to the following 
birth or censoring. 
For estimation purposes, the spells begin nine months after the previous birth because 
this is the earliest we should expect to observe a new birth. 
Censoring can happen for three reasons: The survey takes place, sterilization of the 
woman or her husband, or imposed because of too few births for the method to work.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either birth or censoring of the spell 
occurs.
The time of censoring is assumed to be independent of the hazard rate,
as is standard in the literature.
The two exit states are the birth of a boy, $j=1$, or a girl, $j=2$.

The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2.
 \label{eq:hazard}
\end{equation}
$D_{j}(t)$ is the piece-wise constant baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.
$\mathbf{Z}$ is the nonproportional part, which includes the interactions between 
$D_j(t)$ and a set of explanatory variables and the interactions of those.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.

% To further minimize any potential bias from assuming proportionality, estimations 
% are done separately for different levels of mothers' education and different periods.
% I discuss the choice of which variables to use for nonproportionality below.

Equation \ref{eq:hazard} is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, splitting spells into smaller intervals---here equal to three months---and 
treating them as observations, I can estimate the model using a standard multinomial 
logit model. 
% In the reorganized data, the outcome variable is 0 if the woman does not have a child 
% in a given interval (the base outcome), 1 if she gives birth to a son in that interval, 
% and 2 if she gives birth to a daughter in that interval.

I use the model to predict birth interval lengths, parity progression probabilities, 
and the sex ratio rather than present coefficients because the interpretation 
of competing risk model coefficients is challenging \citep{thomas96}.
The predicted parity progression probability is the likelihood of giving birth by the
imposed censoring based on standard survival curve calculations averaged across all
women in a given sample.

For birth interval lengths, I estimate a set of percentile durations. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% $(100-\frac{80}{2})$ 
mark on her survival curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

% The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
% parity progression probabilities as weights.
% For an individual woman, the estimated percent of births that are boys at $t$, $\hat{Y_t}$, is:  
% \begin{equation}
% \hat{Y}_t 
% = 
% \frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
% { P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
% \times 100,
% \label{eq:probability_son}
% \end{equation}
% where $b$ and $g$ are the predicted probabilities of having a boy and a girl, respectively,
% conditional on the explanatory variables and not having had a child before $t$.
% Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
% $t$ gives the predicted percent of boys over the entire spell for an individual.%
% \footnote{
% Imagine $T=2$. 
% If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
% then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
% }

The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
parity progression probabilities as weights.
To find the individual sex ratio, I estimate the percentage of births that are boys at $t$, 
conditional on not having had a child before $t$.
Weighting the percent boys with the likelihood of exiting the spell with a birth across all 
$t$ gives the predicted percentage of boys over the entire spell for an individual.%
\footnote{
Imagine $T=2$. 
If 54\% and 66\% of births are boys and the likelihood of giving birth 20\% and 40\%, 
then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$\% boys. 
}

\section{Data\label{sec:data}}

The data come from the four rounds of the NFHS
collected in 1992--1993, 1998--1999, 2005--2006, and 2015--2016.
The surveys are large: 89,777, 90,303, 124,385, and 699,686 women, respectively. 
NFHS-1 and NFHS-2 surveyed only ever-married women, while the two later surveys included 
never-married.

I first discuss sample restrictions and their motivations, then introduce the potential 
issue of underreporting of female births and the imposed censoring of birth intervals.
Next, I discuss how the analyses are split by periods to capture the introduction
and changing legality in prenatal sex determination.
Finally, I introduce the explanatory variables, split by whether they enter as
proportional or nonproportional variables.

I focus on the three spells starting from the first birth and ending with the 
fourth birth.
I exclude the interval from marriage to the first birth because many are imputed and the
higher-order intervals because few women had five or more births, especially among the 
better-educated.

I restrict the sample to Hindus for two reasons.
First, Hindus are the majority population group, about 80\% of India’s population.
Second, the prior literature shows that son preference and sex selection vary 
substantially between Hindus and the second largest group, Muslims. 
Combining them and assuming that the baseline hazard is the same would lead to biased 
results.
Because of space constraints and the relatively small number of observations once split 
by education and periods, I do not provide separate results for Muslims or any  
remaining groups, such as Sikhs, Jains, and Christians.

Finally, I exclude visitors and women in any of the following categories: Never married; 
no gauna yet; married more than once; divorced; not living with husband; inconsistent age 
at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported giving birth 
before age 12, had a birth before marriage, or had an interval between births of less than 
nine months.

In addition to the large number of women surveyed and the long period covered, a significant 
benefit of the NFHS over other surveys is that enumerators pay careful attention to the 
spacing between births and probe for ``missed'' births.
For India, the main concern is the underreporting of deceased children, especially a systematic 
recall error where respondents' likelihood of reporting the birth of a deceased child depends 
on the sex of that child.
Unreported deceased children inflate the length of birth intervals and, with declining mortality, 
make changes over time appear too small.
In the online appendix, I provide a detailed analysis of systematic recall error, which 
shows that recall error depends heavily on how long ago a woman was married. 
I, consequently, drop women married 22 years or more.%
\footnote{
Recall error is likely behind the designation of the first two rounds of NFHS as ``moderate 
quality'' in an analysis of the quality of birth histories in DHS surveys and its impact 
on fertility estimates \citep{Schoumaker2014}.
}

To ensure that there are enough births for the method to work, I censor spells at 96 
months (eight years) after a woman can first give birth, equivalent to 105 months 
after the birth of the prior child.
Less than 1\% of observed births occur after the cutoff.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

% %
% \footnote{
% \input{../tables/num_missed.tex}
% }

Direct information on sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--1983, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
% Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2019}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

I use four periods: 1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The first covers the period before sex selection became available and the second from when sex 
selection became available until the Prenatal Diagnostic Techniques Act.
I have split the period from 1995 until 2016 into two to examine if there was support for 
the prior literature's hypothesized reversal in child sex ratios and son 
preference in India.

The allocation of spells into periods is determined by when conception, and, therefore, 
decisions on sex selection can begin. 
Hence, some spells cover two periods, which may bias downward the differences between the 
periods.
Most sterilizations take place soon after giving birth. 
These spells, therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on prior children's sex composition; the 
fewer boys, the lower the probability of sterilization. 
The effect is to bias downward the differences in parity progression probabilities.

I divide the explanatory variables into two groups, nonproportional and proportional. 
The first group consists of characteristics shown in the prior literature to affect the 
spacing choice and the use of sex selection: Mother’s education, sex composition of 
previous children, and area of residence.
To minimize any potential bias from including proportional variables, I estimate separate 
models for each birth interval, education group, and period combination, rather than 
including education as a variable.

I divide education levels into four groups: No education, 1--7 years, 8--11 years, and 
12 and more years.
The latter two correspond to having completed primary and secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and 11 and 12 for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and 5--7 years of schooling completed and the 8--9 
and 10--11 years of schooling completed to ensure sufficient cell sizes.

I capture sex composition with dummy variables for the possible combinations, ignoring 
the ordering of births.%
\footnote{
With sex selection, the composition of prior children is, in principle, endogenous. 
It is beyond the scope of this paper to develop a method for dealing with this issue. 
}
Area of residence is a dummy variable for living in an urban area.

The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the mother's age when the spell begins, the household's
land ownership, and whether it belongs to a scheduled tribe or caste.
Appendix Tables \ref{tab:des_stat1} and \ref{tab:des_stat2} present the descriptive 
statistics.

\section{How Birth Spacing Changed\label{sec:results}}

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
I begin with a broad outline of how parity progression probabilities and
sex ratios have changed over time and by groups.
I then use these results to separately discuss the changes in birth spacing for 
women who do not appear to use sex selection and for those who do.

Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentile birth interval durations in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with 12 or more years of 
education, and the fourth spell for women with 12 or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2\% boys
\citep{jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors and tests for statistically
significant differences across sex compositions are available in the online appendix.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_rural}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for rural women with no education by spell, sex composition, and period}
\label{fig:spacing_low_rural}
\end{figure}

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_urban}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for urban women with no education by spell, sex composition, and period}
% \label{fig:spacing_low_urban}
% \end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_rural}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for rural women with 1--7 years of education by spell, sex composition, and period}
\label{fig:spacing_med_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_urban}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for urban women with 1--7 years of education by spell, sex composition, and period}
\label{fig:spacing_med_urban}
\end{figure}

>>>>

\newcommand{\mco}[1]{\multicolumn{1}{c}{#1}}
\newcommand{\mct}[1]{\multicolumn{2}{c}{#1}}
\newcommand{\mcth}[1]{\multicolumn{3}{c}{#1}}
\newcommand{\X}{$\times$ }
\newcommand{\hs}{\hspace{15pt}}

\renewcommand\floatpagefraction{.9}
\renewcommand\topfraction{.9}
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
Fourth Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
Text and code for this paper are available at 
\href{https://github.com/cportner/sexSelectionSpacing}{\texttt{https://github.com/cportner/sexSelectionSpacing}}.
}
}

@@

\begin{document}
\graphicspath{{./figures/}}
\DeclareGraphicsExtensions{.eps,.jpg,.pdf,.mps,.png}

@@

@@

@@

\begin{figure*}[htpb]
\centering
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
% \caption{Distribution of education by cohort for women 20 years or older
% using the four rounds of the NFHS}
% \label{fig:education_over_time}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_rural}
% \caption{Percentile birth interval durations, sex ratios, and parity progression  
% for rural women with no education by spell, sex composition, and period}
% \label{fig:spacing_low_rural}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_rural}
% \caption{Percentile birth interval durations, sex ratios, and parity progression  
% for rural women with 1--7 years of education by spell, sex composition, and period}
% \label{fig:spacing_med_rural}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_urban}
% \caption{Percentile birth interval durations, sex ratios, and parity progression  
% for urban women with 1--7 years of education by spell, sex composition, and period}
% \label{fig:spacing_med_urban}
\end{figure*}

\begin{figure*}
\centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_rural}
% \caption{Percentile birth interval durations, sex ratios, and parity progression  
% for rural women with 8--11 years of education by spell, sex composition, and period}
% \label{fig:spacing_high_rural}
\end{figure*}
